Strict Pure Functional Programming and Automatic Parallelization\\ (New-project-in-progress Paper)

### 2016-03-24 pubs/tfp16/paper.tex
adds (partial) mini-Haskell summary

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (New-project-in-progess Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straighforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
\item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow2004};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

???give a proper grammar and summarize any tricky translations???

@@

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{color} % for parenthetical notes while paper is in progress
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (New-project-in-progress Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow2004};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

@@

{\color{red} I didn't want to touch the above text, but there's obviously some 
redundancy in the content here...}

Writing STG in the concrete syntax conceived for our project is tedious and
prone to errors.  Compared to Haskell, the eventual source language for the
compiler, STG is relatively verbose, requiring explicit block and expression
delimiters, and classification of heap objects. Though its general structure is
similar to primitive Haskell, it lacks many of the conveniences that make
Haskell a pleasure to work with.  The motivation for a more expressive language
for testing our compiler in its nascent stages is then clear.  This language,
mini-Haskell, was progressively "grown" from the STG syntax towards a subset of
Haskell itself*, removing much of the frustrating syntactic cruft and adding
some of the core features that are often taken for granted in Haskell (and other
languages).

{\color{red}
* Exceptions include the "unboxed" keyword allowed in datatype definitions and
the ability to specify unboxed ints and doubles with a postfix "\#" character.
The former feature is not in any extension of Haskell, while the latter can be
included using the MagicHash extension.}

An informal specification of the mini-Haskell grammar is given in
{\color{red}numbered table?}.  There is significant overlap with STG, resulting
from the structural similarities, but the redundancies are included for
completeness. {\color{red} should they be?}

% Haskell Report:
%   https://www.haskell.org/onlinereport/haskell2010/haskell.html
% (PDF):
%   https://www.haskell.org/definition/haskell2010.pdf

% Chapter 10:
%   https://www.haskell.org/onlinereport/haskell2010/haskellch10.html
% (No standalone PDF)

% because # looks strangely large next to math terms
\newcommand{\hash}{{\scriptsize\#}}

\begin{table}
\centering
\begin{tabular}{r r c l l}
Variable         & $f,\ x$             &     &
& Initial lower-case letter \\

Constructor        & $C$                 &     & 
& Initial upper-case letter \\

Literal            & $\mathit{lit}$      & ::= & $i\ |\ d\ 
                                                 |\ i$\hash\ $|\ d$\hash
& Integral or floating point literal (boxed) \\

                   &                     &     & 
& optional \# suffix denotes unboxed value \\

Atom               & $a$                 & ::= & $\mathit{lit}\ |\ x\ |\ C$
& Constructors are first class \\ 
% not sure if that's the right terminology 
\\

Expression         & $e$                 & ::= & $a$
& Atomic expression \\

                   &                     & $|$ & $e\ e$
& Expression application \\

                   &                     & $|$ & \texttt{\textbackslash} $\mathit{pat}_i$ 
                                               \texttt{->} $e$
& Lambda expression, $i > 0$ \\

                   &                     & $|$ & \texttt{case} $e$ \texttt{of} $\mathit{alt}_i$
& Case expression, $i > 0$\\

                   &                     & $|$ & \texttt{let} $\mathit{odecl}_i$ \texttt{in} $e$
& Recursive let expression, $i > 0$ \\

\\

Pattern            & $\mathit{pat}$      & ::= & $C\ \mathit{pat}_i$
& Nested constructor matching \\
                   &                     & $|$ & $\mathit{lit}$
& Match numeric literals \\
                   &                     & $|$ & $x$
& Bind to variable \\

\\

Alternative        & $\mathit{alt}$      & ::= & $\mathit{pat}_i$ \texttt{->} $e$
& $i > 0$ \\

\\

Type               & $\mathit{type}$     & $|$ & $C$
& Concrete type (constructor naming) \\
                   &                     & $|$ & $x$
& polymorphic type variable \\
                   &                     & $|$ & $\mathit{type}$ \texttt{->} $\mathit{type}$
& Function type \\

\\

Object Declaration & $\mathit{odecl}$    & ::= & $x$ \texttt{::} $\mathit{type}$
& Type signature \\
                   &                     & $|$ & $x = e$
& Simple binding \\
                   &                     & $|$ & $f\ \mathit{pat}_i = e$
& Function declaration \\
                   &                     &     &
& $e$ cannot be an unboxed literal \\

\\

Constructor Definition & $\mathit{con}$  & ::= & $C\ \mathit{type}_i$
& $i \ge 0$ \\

Datatype Declaration &  $\mathit{ddecl}$ & ::= & \texttt{data} 
                                                 [\texttt{unboxed}]
                                                 $C\ x_i =$ 
& User-defined data type, $i \ge 1, n > 0$ \\
                     &                   &     & $\mathit{con}_1 | \dots |
                                                 \mathit{con}_n$
& \\

\\

Program                & $\mathit{prog}$ & ::= & $(o|d)\mathit{decl}_i$
& Objects and datatypes, no order necessary

Because mini-Haskell was designed to replace STG, it makes sense to talk about
it in terms of the differences between the two and how mini-Haskell is
translated into STG for the backend.  Here, the differences are given in order
of perceived utility, or importance to someone transition from STG to
mini-Haskell.

\paragraph{Removal of explicit heap objects} -\\

In STG, all object definitions require explicit classification with
\texttt{CON}, \texttt{PAP}, \texttt{THUNK}, \texttt{FUN}, or
\texttt{BLACKHOLE}. This classification can be inferred from context by the
compiler, so these keywords are no longer necessary.  Functions can then be
defined in Haskell-style, with the parameters following the function's
identifier on the left side of the equality. With these simple changes, the
syntax becomes a subset of Haskell's*.

Heap objects are still only created at the top level or in \texttt{let}
expressions, so the translation is straightforward: \texttt{FUN} objects are
identifiable by the inclusion of these parameters.  \texttt{PAP} objects are
created when a known function is partially applied.  \texttt{CON} objects are
created when constructors are fully applied.  Every other object becomes a
\texttt{THUNK}.

{\color{red}
* With only these changes, the semantics are a little different; numeric
  literals would still be treated as unboxed values, for example.  I don't know
  if that's worth noting though.}

\paragraph{The Layout Rule} -\\

Mini-Haskell implements Haskell's Layout Rule in an almost direct translation of
the specification given in section 10.3 of the Haskell 2010 Report.  The notable
exception is the omission of any attempt to define the \texttt{parse-error}
function.  This is apparently a tricky spot for Haskell compilers (even GHC has
not always had it right) so, in the name of tradition, it remains a tricky spot
for our mini-Haskell front end.  The edge cases that it will incorrectly handle
as a result of this are exactly that: edge cases.  As such, they are easy to
avoid in practical usage.

The inclusion of the layout rule may seem small and insignificant, but it allows
its users to produce code that is visually similar to Haskell code.  This makes
it easier to read and write for any Haskell programmer who relies on layout to
delineate context (I assume this is ~99.9% of human Haskell users).  It also
allows for the use of Haskell programming environments that support "smart"
indentation.

% Layout Rule Spec:
%   https://www.haskell.org/onlinereport/haskell2010/haskellch10.html\#x17-17800010.3
% GHC's syntax "infelicities":
%   https://downloads.haskell.org/~ghc/latest/docs/html/users\_guide/bugs-and-infelicities.html\#infelicities-syntax
% More on parse-error:
%   http://www.dcs.gla.ac.uk/mail-www/haskell/msg01711.html

\paragraph{Expression to Expression application} -\\

In STG, function application must be to atomic values, i.e. either literals or
variables.  This often requires introducing bindings with either \texttt{let} or
\texttt{case} expressions.  Consider the \texttt{append} function that
concatenates its two list parameters.  In STG, it is written as:

\begin{verbatim}
append = FUN(l1 l2 ->
            case l1 of
              { Nil -> l2;
                Cons hd tl -> 
                  let { rec = THUNK(append tl l2);
                        result = CON(Cons hd rec) }
                  in result });
\end{verbatim}

In mini-Haskell, expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:

\begin{verbatim}
append l1 l2 = case l1 of
                Nil -> l2
                Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}

This also serves as a good example of expression to expression application is
transformed into STG.  Any non-atomic expression in a expression application is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly "atomized" for STG.  The above mini-Haskell code
is transformed in STG that is structurally identical to the STG code that
precedes it.

\paragraph{Fancy Pattern Matching} -\\

{\color{red}\dots}

\paragraph{Less significant features} -\\

\begin{itemize}
\item Constructors as functions \\
  (partial application uses generated functions, full creates CONS)

\item Primitive operations as functions \\
  (partial application uses generated functions, full creates EPrimop)

\item Lambda Expressions \\
  (desugar to let-bound FUN objects)

\item Type Annotations \\
  (passed on to type checker)

\item Haskell Block \\
  (ignored by MHS frontend, not other Haskell compilers;
  allows the same file to be tested on our compiler vs GHC)
\end{itemize}
